---
title: "07 Non-Proportional Substitution Patterns II"
output: html_notebook
---

# Non-Proportional Substitution Patterns I: The Probit Model {#chapter-7}

> "Perfectionism means that you try not to leave so much mess to clean up."
>
> --- Anne Lamott, Bird by Bird: Some Instructions on Writing and Life

> "Perfection’s unattainable but it isn’t unapproachable."
>
> --- Peter Watts, Blindsight

## Preliminaries

Load the packages used in this section:
```{r}
library(tidyverse)
library(mlogit)
library(kableExtra)
```

Load the dataset used in this section (from the `mlogit` package):
```{r}
data("Heating")
```

The dataset is in "wide" form, which means that there is one record per decision making unit (i.e. per household), so we need to change the data to "long" format:
```{r}
H <- mlogit.data(Heating, shape = "wide", choice = "depvar", varying = c(3:12))
```

This is the same dataset used in the preceding chapter with five heating systems:

- Gas Central (gc)
- Gas Room (gr)
- Electric Central (ec)
- Electric Room (er)
- Heat Pump (hp)

This dataset was used to illustrate the nested logit model. The results before suggested that there was a strong correlation among central systems and room-based systems, and therefore a nested logit model was an improvement over the multinomial logit model.

## More flexible substitution patterns

**NOTE** This chapter is work in progress and is not yet complete.

In Chapter \@ref(chapter-6) the topic of non-proportional substitution was discussed, and a method for deriving logit models using the Generalized Extreme Value method was presented. In this chapter, an alternative model for more flexible substitution patterns is introduced. This model is based on the use of the normal distribution for the random utility, instead of the extreme value distribution. The model is called _multinomial probit_ and can be used to represent arbitrary substitution patterns. On the downside, the model is technically and computationally more demanding.

## How to use this note

Remember that the source for the document you are reading is an R Notebook. Throughout the notes, you will find examples of code in segments of text called _chunks_. This is an example of a chunk:
```{r}
print("If you are always trying to be normal, you will never know how amazing you can be.")
```

If you are working with the Notebook version of the document, you can run the code by clicking the 'play' icon on the top right corner of the chunk. If you are reading the web-book version of the document, you will often see that the code has already been executed. You can still try it by copying and pasting into your R or RStudio console.

## Learning objectives

In this practice, you will learn about:

1. The probit model.
2. Identification of parameters.
3. Estimation of the model.

## Suggested readings

- Ben-Akiva, M. Lerman, [-@Benakiva1985discrete] Discrete Choice Analysis: Theory and Applications to Travel Demand, **Chapter 5, pp. 128-129**, MIT Press.
- Ortuzar JD, Willumsen LG [-@Ortuzar2011modelling] Modelling Transport, Fourth Edition, **Chapter 7, pp. 248-250**, John Wiley and Sons.
- Train [-@Train2009discrete] Discrete Choice Methods with Simulation, Second Edition, **Chapter 5**, Cambridge University Press.

## Probit model

Assume that the choice problem of decision-maker $n$ consists of $J$ alternatives, and that the utility of each alternative is decomposed in two parts, a systematic utility $V$ and a random utility $\epsilon$, as follows:
$$
U_{n1} = V_{n1} + \epsilon_{n1}\\
\cdots\\
U_{nj} = V_{nj} + \epsilon_{nj}\\
\cdots\\
U_{nJ} = V_{nJ} + \epsilon_{nJ}
$$
The random utility terms can be collected as follows:
$$
\epsilon_n = [\epsilon_1, \cdots, \epsilon_j, \cdots, \epsilon_J] 
$$

Previously, the logit model was derived based on the assumption that the random utility terms $\epsilon_j$ were Extreme Value Type I. Now, the assumption is that the vector of random utilities $\epsilon_n$ follows the normal distribution with a mean vector of zero and a covariance matrix $\Sigma_n$ that can be a function of variables specific to a decision-maker. The probability distribution function is:
$$
\phi(\epsilon_n)=\frac{1}{(2\pi)^{J/2}|\Sigma_n|^{1/2}}e^{-\frac{1}{2}\epsilon_n^T\Sigma_n^{-1}\epsilon_n}
$$

As before, the probability of choosing alternative $j$ is given by the following expression:
$$
P_j=P(U_j\ge U_k;\forall k\ne j)
$$

Since the probability is a function of the difference in utilities, it is possible to define the following expressions for differences with respect to the utility of the alternative whose probability we wish to calculate (i.e., $j$):
$$
\tilde{U}_{nkj}=U_{nk} - U_{nj}
$$

which implies:
$$
\tilde{V}_{nkj}=V_{nk} - V_{nj}\\
\tilde{\epsilon}_{nkj} = \epsilon_{nk} - \epsilon_{nj}
$$

In this case, the probability expression becomes:
$$
P_j=P(U_{nkj}\le 0;\forall k\ne j)
$$


$$
\phi(\tilde{\epsilon}_n)=\frac{1}{(2\pi)^{J/2}|\tilde{\Sigma}_n|^{1/2}}e^{-\frac{1}{2}\tilde{\epsilon}_n^T\tilde{\Sigma}_n^{-1}\tilde{\epsilon}_n}
$$



Now we are ready to estimate a multinomial logit model (we called this Model 3 in Chapter \@ref(chapter-3)):
```{r}
mod3 <- mlogit(depvar ~ ic + oc, H, reflevel = "ec")
summary(mod3)
```

```{r}
nl2 <- mlogit(depvar ~ ic + oc, H,
             nests = list(room = c( 'er', 'gr'), central = c('ec', 'gc', 'hp')),
             un.nest.el = TRUE, steptol = 1e-12)
summary(nl2)
```

```{r eval=FALSE}
prob1 <- mlogit(depvar ~ ic + oc, H, reflevel = "ec", probit = TRUE, R=100)
summary(prob1)
```

```{r eval=FALSE}
prob2 <- mlogit(depvar ~ ic + oc, H, reflevel = "ec", probit = TRUE,
                constPar=c('hp.hp', "gr.hp", "gr.gr", "gc.gr", "gc.gc", "er.hp"))
summary(prob2)
```

```{r eval=FALSE}
prob3 <- mlogit(depvar ~ ic + oc, H, reflevel = "ec", probit = TRUE,
                constPar=c("er.er", "gr.hp", "gc.gr", "er.hp", "er.gc"), R = 50)
summary(prob3)
```

```{r eval=FALSE}
L1 <- matrix(0, 4, 4)
L1[!upper.tri(L1)] <- c(1, 0, coef(prob3)[7], 0, coef(prob3)[8], 0, coef(prob3)[9:10], 0, coef(prob3)[11])
L1
```

```{r eval=FALSE}
L1 %*% t(L1)
```

```{r eval=FALSE}
L1 <- matrix(0, 4, 4)
L1[!upper.tri(L1)] <- c(1, coef(prob1)[7:15])
L1
```

```{r eval=FALSE}
L1 %*% t(L1)
```

To simulate this scenario, we begin by copying the input dataframe:
```{r eval=FALSE}
H_rebate <- H
```

In the new dataframe that will simulate the rebate, replacing the cost of installation as follows: 
```{r eval=FALSE}
H_rebate[H_rebate$alt == "gc", "ic"] <- 1.15 * H_rebate[H_rebate$alt == "gc", "ic"]
```

We can calculate the market shares of the "do nothing" and "rebate" policies and compare their shares (which are the mean values of the predictions):
```{r eval=FALSE}
data.frame(Policy = c("Do nothing", "15% rebate"),
           rbind(apply(predict(mod3, newdata = H), 2, mean),
                 apply(predict(mod3, newdata = H_rebate), 2, mean)))
```

Nested logit:
```{r eval=FALSE}
data.frame(Policy = c("Do nothing", "15% rebate"),
           rbind(apply(predict(nl2, newdata = H), 2, mean),
                 apply(predict(nl2, newdata = H_rebate), 2, mean)))
```

Probit:
```{r eval=FALSE}
data.frame(Policy = c("Do nothing", "15% rebate"),
           rbind(apply(predict(prob3, newdata = H), 2, mean),
                 apply(predict(prob3, newdata = H_rebate), 2, mean)))
```

## Exercise

1. 