---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# A reproducible environment for the book Discrete Choice Analysis with R

<!-- badges: start -->

<!-- badges: end -->

```{r, include=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

This repository makes use of package
{[renv](https://rstudio.github.io/renv/)} to provide a reproducible
project environment to work with the book [Discrete Choice Analysis with
R](https://doi.org/10.1007/978-3-031-20719-8).

The repository provides all the infrastructure to replicate the
environment used to create the book. This includes the version of `R` as
well as all packages used in the book.

## How to use this repository

1.  Install R (make sure to pick the correct operating system) - this is
    the programming language, it comes as a 'core package' here:
    
    <https://cran.rstudio.com/>

    Currently, the environment runs with R 4.4.2 (Pile of Leaves). You can get it
    here:

    <https://www.r-project.org>

3.  Install RStudio an Interactive Development Environment (IDE) for `R`. It is
    available here:
    
    <https://www.rstudio.com/products/rstudio/download/>.

4.  Download the Rtools44 installer (Windows) and run it to install:

    <https://cran.r-project.org/bin/windows/Rtools/rtools44/rtools.html>.
    
    If using a Mac this may be achieved through installing Xcode *untested*:
    <https://mac.r-project.org/tools/>

5.  Download the reproducible environment as a .zip file from this repository:
    <https://github.com/paezha/Discrete-Choice-Analysis-with-R>. To do so, right-click
    on the "code" button on the top-right corner and select "Download ZIP". 
    
    The repository contains a `renv.lock` file that specifies all the packages 
    used in the [textbook](https://doi.org/10.1007/978-3-031-20719-8). 
    Unzip the file and store it in an appropriate directory. This becomes your
    project folder that you can use  to work with this book.

6.  Navigate to your project folder and double click the `.RProj` file called
    "Discrete-Choice-Analysis-with-R". This will launch RStudio.

7.  Install your first `R` package {renv}

<!-- -->

    install.packages("renv")

7.  Close Rstudio. Restart your computer.

8.  Double click the Discrete-Choice-Analysis-with-R.Rproj file. This will
    relaunch RStudio. You will see a message in your console saying that
    your library is out of sync with the lock file. Let's make sure
    they are both in sync by 'restoring' the state of your computer to
    match the versions specified in the .lock file. Run the following:

<!-- -->

    renv::restore()

If an individual library does not install, run:

    install.packages("XXXXX") #where the XXXXX are replaced with the package name
    #and then run again..
    renv::restore()

If this issue occurs again, repeat the above chunk.

9.  Restart `R` by going to Session -\> Restart R. Now click the "+"
    green button and open a new "R Markdown" file. Select the 4th option
    in the left panel "From Template". Select a template option from the
    [{discrtr}](https://paezha.github.io/discrtr/) package. Each template is one
    chapter in the book. Give the template a name, say "my-template-chapter-1".

10. Knit the "my-template-chapter-1.Rmd" to a .html file by clicking the
    arrow on the "Knit" button.

You’ve just knitted your first chapter! Congrats.

Feel free to use the .Rmd files as your notebooks: annotate them, write your 
comments, thoughts, and questions, modify the code and experiment. When you 
finish all the templates you will have a complete set of personal notes on 
the topics of the book.
